Correcting for sample selection in Bayesian analyses

### 2017-05-26 papers/general_method/general.tex
More feedback from Alex

With this sample selection, all events satisfy $x > \alpha$, giving $P(S|x,\theta) = \mathcal{H}(x - \alpha)$. We assign a value $\alpha=85$ for convenience. If we do not take this truncation into account, we will recover biased parameter estimates. However, we can correct for this truncation using equation \eqref{eq:main}, as the integral in the denominator has an analytic solution. Having successfully observed $x_i$, it follows that $x_i > \alpha$ and so $P(S|x_i,\theta) = 1$. To substitute in our normal model,
\begin{align}
\mathcal{L}_i &= \frac{P(S|x_i,\theta) P(x_i|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD} \\
&= \frac{ \mathcal{N}(x_i|\mu, \sigma)}{\int_{-\infty}^\infty \mathcal{H}(D - \alpha) \mathcal{N}(D|\mu, \sigma)\, dD} \\
&= \frac{ \mathcal{N}(x_i|\mu, \sigma)}{\int_{\alpha}^\infty \mathcal{N}(D|\mu, \sigma)\, dD} \\
&= \frac{ \mathcal{N}(x_i|\mu, \sigma)}{\frac{1}{2} {\rm erfc}\left[ \frac{\alpha - \mu}{\sqrt{2}\sigma} \right]}, 
\end{align}
where in the last line we have evaluated the integral in the case $\mu > \alpha$. Note that this is for a single observation, and so for a set of independent observations we need to introduce the product found in equation \eqref{eq:prod}. 
\begin{align}
\mathcal{L} &= \prod_{i=1}^N \mathcal{L}_i =  \prod_{i=1}^N \frac{ \mathcal{N}(x|\mu, \sigma)}{\frac{1}{2} {\rm erfc}\left[ \frac{\alpha - \mu}{\sqrt{2}\sigma} \right]}, 
\end{align}
However, as our selection efficiency correction is observationally independent it is identical for all observations, allowing us to take it outside the product.
\begin{align}
\mathcal{L} &= 2 \left( {\rm erfc}\left[ \frac{\alpha - \mu}{\sqrt{2}\sigma} \right] \right)^{-N}\prod_{i=1}^N  \mathcal{N}(x|\mu, \sigma), 
\end{align}
We can add this correction to our model, and note that we now recover unbiased parameter estimates. This is demonstrated in Figure \ref{fig:imperfect}, which shows the posterior surfaces for when you take sample selection into account and when you do not.
\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{example/imperfect.pdf}
	\end{center}
	\caption{A systematic test of our imperfect model, done by stacking the output chains from fitting 100 independent realisations of our 100 data points, subject to our thresholding. The bias shown in the red `Biased' contour can be corrected to via the techniques shown in Section \ref{sec:imperfect} to recover unbiased surfaces.}
	\label{fig:imperfect}
\end{figure}

@@

So let us modify our imperfect toy model from the previous section. Instead of observing just one variable, $x$, we also observe a new independent variable, $y$, which is drawn from its own distribution $y \sim \mathcal{N}(\mu_y, \sigma_y)$. Our selection efficiency can now become a combination of $x$ and $y$, such that we only observe events that satisfy $x + \beta y > \alpha$, giving $P(S|x,y,\theta) = \mathcal{H}(x + \beta y - \alpha)$.  Our likelihood for such a toy model becomes now the combination of probabilities for observing both $x$ and $y$, with the denominator becoming an integral over all possible $X$ and $Y$ observations subject to our selection effects.
\begin{align}
\mathcal{L}_i &= \frac{ \mathcal{N}(x_i|\mu, \sigma) \mathcal{N}(y_i|\mu_y, \sigma_y)}
{\iint_{-\infty}^\infty \mathcal{H}(X + \beta Y - \alpha) \mathcal{N}(X|\mu, \sigma) \mathcal{N}(Y|\mu_y, \sigma_y)\, dX dY}
\end{align}
Assume that we cannot solve this integral analytically, and must resort to numeric solutions. These often clash with sampling methods, especially for high dimensional integrals. Inserting Monte Carlo integration into fitting algorithms can drastically slow them down, and algorithms such as Hamiltonian MCMC that require continuous surfaces can easily fail on surfaces that fluctuate from Monte Carlo integration. Even by fixing the samples used in MC integration (thereby giving a continuous surface), the complexity of the surface derivatives will pose almost insurmountable problems for any algorithms that utilise surface gradients. One solution is to find an approximate, analytic correction we can utilise in our fitting algorithm which seeks to shift the region of parameter space sampled by the sampler closer to the correct area.

In our example, if $\beta \ll 1$, such that the majority of selection effect is encapsulated by $x$ and not $y$, our approximate correction can take the form found in the previous correction from Section \ref{sec:imperfect}. Having true values of $\mu = 100$, $\sigma = 10$, $\mu_y = 30$, $\sigma_y = 5$, and a known $\beta = 0.2$, we can give a concrete example. Assuming some prior, imperfect knowledge of $\mu_y$ (perhaps we believe it is approximately $20$) we estimate that the average contribution from $\beta y$ is approximately $20\beta = 4$ (which is close to the correct value of $6$), and from this our analytic correction to our likelihood can be
\begin{align}
w_{\rm approx} = \frac{1}{2} {\rm erfc}\left[ \frac{\alpha - \mu - 4}{\sqrt{2}\sigma} \right].
\end{align}
Further, let us explicitly break our likelihood into two parts, $\mathcal{L}_i = \mathcal{L}_{i1} \mathcal{L}_{i2}$, with the parts given by
\begin{align}
\mathcal{L}_{i1} &= \frac{ \mathcal{N}(x_i|\mu, \sigma) \mathcal{N}(y_i|\mu_y, \sigma_y)}{w_{\rm approx} } \\
\mathcal{L}_{i2} &= \frac{ w_{\rm approx}  }{\iint_{-\infty}^\infty \mathcal{H}(x + \beta y - \alpha) \mathcal{N}(X|\mu, \sigma) \mathcal{N}(Y|\mu_y, \sigma_y)\, dX dY}.
\end{align}
$\mathcal{L}_1$ can thus be fitted with a traditional sampler without numeric difficulty or slowdown, and $\mathcal{L}_2$ allows us to calculate the weight of each sample. We are effectively importance sampling our likelihood evaluations. The computational benefits of this should not be understated either - each sample in our chains can be reweighted independently, providing a task that is trivially parallelisable. Evaluating $\mathcal{L}_2$ using Monte Carlo integration of $n$ samples, we have
\begin{align}
\mathcal{L}_{i2} = \frac{ w_{\rm approx}  n }{\sum_{j=1}^{n} \mathcal{H}(X_j + \beta Y_j - \alpha) \mathcal{N}(X_j|\mu, \sigma) \mathcal{N}(Y_j|\mu_y, \sigma_y)} \label{eq:mc}.
\end{align}
We can now easily move from a single observation to a set of $N$ observations.
\begin{align}
\mathcal{L}_1 &= w_{\rm approx}^{-N} \prod_{i=1}^N \mathcal{N}(x_i|\mu, \sigma) \mathcal{N}(y_i|\mu_y, \sigma_y) \\
\mathcal{L}_{2} &= \left(  \frac{ w_{\rm approx}  n }{\sum_{j=1}^{n} \mathcal{H}(X_j + \beta Y_j - \alpha) \mathcal{N}(X_j|\mu, \sigma) \mathcal{N}(Y_j|\mu_y, \sigma_y)}\right)^N
\end{align}

\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{example/real.pdf}
	\end{center}
	\caption{A systematic test of our more complicated model, done by stacking the output chains from fitting 100 independent realisations of our 100 data points, subject to our thresholding. The likelihood $\mathcal{L}_1$ was evaluated, and reweighted using Monte Carlo integration of a hundred thousand possible events as per $\mathcal{L}_2$. The truncated data with no correction is shown as `Biased' in dotted red, the `Approximate` only correction ($\mathcal{L}_1$) shown in dashed green, and the final reweighted chain shown in solid black as `Corrected'.}
	\label{fig:real}
\end{figure}
Thus we end up with a corrected posterior surface as shown in Figure \ref{fig:real}.

@@

\subsection{Importance Sampling}
Further tricks can be used to increase the efficiency with which the samples are reweighted. Firstly, the overarching analytic model often provides functions which can be drawn from efficiently. In the case of our example, by drawing the random numbers $X$ and $Y$ respectively from the normal distributions $\mathcal{N}(\mu,\sigma)$ and $\mathcal{N}(\mu_y,\sigma_y)$ (ie traditional importance sampling) we need only evaluate the step function for our data points.

>>>>

With this sample selection, all events satisfy $x > \alpha$, giving $P(S|x,\theta) = \mathcal{H}(x - \alpha)$. We assign a value $\alpha=85$ for convenience. If we do not take this truncation into account, we will recover biased parameter estimates. However, we can correct for this truncation using equation \eqref{eq:main}, as the integral in the denominator has an analytic solution. Having successfully observed $x_i$, it follows that $x_i > \alpha$ and so $P(S|x_i,\theta) = 1$. To substitute in our normal model,
\begin{align}
\mathcal{L}_i &= \frac{P(S|x_i,\theta) P(x_i|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD} \\
&= \frac{ \mathcal{N}(x_i|\mu, \sigma)}{\int_{-\infty}^\infty \mathcal{H}(D - \alpha) \mathcal{N}(D|\mu, \sigma)\, dD} \\
&= \frac{ \mathcal{N}(x_i|\mu, \sigma)}{\int_{\alpha}^\infty \mathcal{N}(D|\mu, \sigma)\, dD} \\
&= \frac{ \mathcal{N}(x_i|\mu, \sigma)}{\frac{1}{2} {\rm erfc}\left[ \frac{\alpha - \mu}{\sqrt{2}\sigma} \right]}, 
\end{align}
where in the last line we have evaluated the integral in the case $\mu > \alpha$. Note that this is for a single observation, and so for a set of independent observations we need to introduce the product found in equation \eqref{eq:prod}. 
\begin{align}
\mathcal{L} &= \prod_{i=1}^N \mathcal{L}_i =  \prod_{i=1}^N \frac{ \mathcal{N}(x|\mu, \sigma)}{\frac{1}{2} {\rm erfc}\left[ \frac{\alpha - \mu}{\sqrt{2}\sigma} \right]}, 
\end{align}
However, as our selection efficiency correction is observationally independent it is identical for all observations, allowing us to take it outside the product.
\begin{align}
\mathcal{L} &= 2 \left( {\rm erfc}\left[ \frac{\alpha - \mu}{\sqrt{2}\sigma} \right] \right)^{-N}\prod_{i=1}^N  \mathcal{N}(x|\mu, \sigma), 
\end{align}
We can add this correction to our model, and note that we now recover unbiased parameter estimates. This is demonstrated in Figure \ref{fig:imperfect}, which shows the posterior surfaces for when you take sample selection into account and when you do not. The selection bias preferentially selects intrinsically brighter objects and, by cutting out some of the distribution, narrows the observed distribution, and so we note that the bias correction correctly increases the weight of low $\mu$ and high $\sigma$ parametrisations, as those models would be subject to the most sample bias. We also note that not only does the best fit location for each parameter shift, but the shape of the posterior surface itself is significantly modified.
\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{example/imperfect.pdf}
	\end{center}
	\caption{A systematic test of our imperfect model, done by stacking the output chains from fitting 100 independent realisations of our 100 data points, subject to our thresholding. The bias shown in the red `Biased' contour can be corrected to via the techniques shown in Section \ref{sec:imperfect} to recover unbiased surfaces.}
	\label{fig:imperfect}
\end{figure}

@@

In this subsection we introduce an example that presents computational challenges in the evaluation of the posterior, and which better reflects the model complexity in supernova cosmology analysis. The specific challenge is that the denominator in the equation for the likelihood equation \eqref{eq:main} is dependent on $\theta$ and does not have an analytic solution (i.e. has to be solved numerically). When using Monte Carlo sampling methods to probe the posterior with this kind of likelihood, the integral in the denominator must be calculated with each draw. We therefore present importance sampling as an approach that make these posteriors more computationally tractable.

Let us modify our imperfect toy model from the previous section. Instead of observing just one variable, $x$, we also observe a new independent variable, $y$, which is drawn from its own distribution $y \sim \mathcal{N}(\mu_y, \sigma_y)$, and has no measurement uncertainty (like $x$). Our selection efficiency can now become a combination of $x$ and $y$, such that we only observe events that satisfy $x + \beta y > \alpha$, giving $P(S|x,y,\theta) = \mathcal{H}(x + \beta y - \alpha)$.  Our likelihood for such a toy model becomes now the combination of probabilities for observing both $x$ and $y$, with the denominator becoming an integral over all possible $X$ and $Y$ observations subject to our selection effects.
\begin{align}
\mathcal{L}_i &= \frac{ \mathcal{N}(x_i|\mu, \sigma) \mathcal{N}(y_i|\mu_y, \sigma_y)}
{\iint_{-\infty}^\infty \mathcal{H}(X + \beta Y - \alpha) \mathcal{N}(X|\mu, \sigma) \mathcal{N}(Y|\mu_y, \sigma_y)\, dX dY}
\end{align}
Assume that we cannot solve this integral analytically, and must resort to numeric solutions. These often clash with sampling methods, especially for high dimensional integrals. Inserting Monte Carlo integration into fitting algorithms can drastically slow them down, and algorithms such as Hamiltonian MCMC that require continuous surfaces can easily fail on surfaces that fluctuate from Monte Carlo integration. Even by fixing the samples used in MC integration (thereby giving a continuous surface), the complexity of the surface derivatives will pose almost insurmountable problems for any algorithms that utilise surface gradients.

One solution is to find an approximate, analytic correction we can utilise in our fitting algorithm which seeks to shift the region of parameter space sampled by the sampler closer to the correct area, and then importance sample our MC chains to provide a fully corrected surface. Given an approximate analytic correction $w_{\rm approx}$, we explicitly break our likelihood into two parts, $\mathcal{L}_i = \mathcal{L}_{i1} \mathcal{L}_{i2}$, with the parts given by
\begin{align}
\mathcal{L}_{i1} &= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{w_{\rm approx}} \\
\mathcal{L}_{i2} &= \frac{w_{\rm approx}}{\int P(S, D|\theta)\, dD}
\end{align}
Where $\mathcal{L}_{i1}$ represents an analytic function which can efficiently be sampled using traditional Monte Carlo sampling methods, and $\mathcal{L}_{i2}$ is a numerically calculated weight applied after Monte Carlo sampling in order to model the full likelihood surface.

In our example, if $\beta \ll 1$, such that the majority of selection effect is encapsulated by $x$ and not $y$, our approximate correction can take the form found in the previous correction from Section \ref{sec:imperfect}. Having true values of $\mu = 100$, $\sigma = 10$, $\mu_y = 30$, $\sigma_y = 5$, and a known $\beta = 0.2$, we can give a concrete example. Assuming some prior, imperfect knowledge of $\mu_y$ (perhaps we believe it is approximately $20$) we estimate that the average contribution from $\beta y$ is approximately $20\beta = 4$ (which is close to the correct value of $6$), and from this we add in a small adjustment to the analytic correction from Section \ref{sec:imperfect}:
\begin{align}
w_{\rm approx} = \frac{1}{2} {\rm erfc}\left[ \frac{\alpha - \mu - 4}{\sqrt{2}\sigma} \right].
\end{align}
This gives our likelihood parts as
\begin{align}
\mathcal{L}_{i1} &= \frac{ \mathcal{N}(x_i|\mu, \sigma) \mathcal{N}(y_i|\mu_y, \sigma_y)}{w_{\rm approx} } \\
\mathcal{L}_{i2} &= \frac{ w_{\rm approx}  }{\iint_{-\infty}^\infty \mathcal{H}(x + \beta y - \alpha) \mathcal{N}(X|\mu, \sigma) \mathcal{N}(Y|\mu_y, \sigma_y)\, dX dY}.
\end{align}
As stated previously, $\mathcal{L}_1$ can thus be fitted with a traditional sampler without numeric difficulty or slowdown, and $\mathcal{L}_2$ allows us to calculate the weight of each sample. We are effectively importance sampling our likelihood evaluations. The computational benefits of this should not be understated either - each sample in our chains can be reweighted independently, providing a task that is trivially parallelisable. Evaluating $\mathcal{L}_2$ using Monte Carlo integration of $n$ samples, we have
\begin{align}
\mathcal{L}_{i2} = \frac{ w_{\rm approx}  n }{\sum_{j=1}^{n} \mathcal{H}(X_j + \beta Y_j - \alpha) \mathcal{N}(X_j|\mu, \sigma) \mathcal{N}(Y_j|\mu_y, \sigma_y)} \label{eq:mc}.
\end{align}
We can now easily move from a single observation to a set of $N$ observations.
\begin{align}
\mathcal{L}_1 &= w_{\rm approx}^{-N} \prod_{i=1}^N \mathcal{N}(x_i|\mu, \sigma) \mathcal{N}(y_i|\mu_y, \sigma_y) \\
\mathcal{L}_{2} &= \left(  \frac{ w_{\rm approx}  n }{\sum_{j=1}^{n} \mathcal{H}(X_j + \beta Y_j - \alpha) \mathcal{N}(X_j|\mu, \sigma) \mathcal{N}(Y_j|\mu_y, \sigma_y)}\right)^N
\end{align}
Thus we end up with a corrected posterior surface as shown in Figure \ref{fig:real}. It is important to note that the point of maximum likelihood is biased by roughly the same amount in the biased and approximate posterior surfaces in Figure \ref{fig:real}. However, as the approximately correctly posterior is broader than the biased surface, it has far more samples in the region of parameter space mapped by the fully corrected posterior. This is the entire purpose of the approximate correction - to maximise the number of samples in the correct region of parameter space, so that we can importance sample our chains efficiently.
\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{example/real.pdf}
	\end{center}
	\caption{A systematic test of our more complicated model, done by stacking the output chains from fitting 100 independent realisations of our 100 data points, subject to our thresholding. The likelihood $\mathcal{L}_1$ was evaluated, and reweighted using Monte Carlo integration of a hundred thousand possible events as per $\mathcal{L}_2$. The truncated data with no correction is shown as `Biased' in dotted red, the `Approximate` only correction ($\mathcal{L}_1$) shown in dashed green, and the final reweighted chain shown in solid black as `Corrected'.}
	\label{fig:real}
\end{figure}

@@

\subsection{Importance Sampling}
Further tricks can be used to increase the efficiency with which the samples are reweighted. Firstly, the overarching analytic model often provides functions which can be drawn from efficiently. In the case of our example, by drawing the random numbers $X$ and $Y$ respectively from the normal distributions $\mathcal{N}(\mu,\sigma)$ and $\mathcal{N}(\mu_y,\sigma_y)$ (ie traditional importance sampling) we need only evaluate the step function for our data points. That is, we replace
\begin{align}
\int f(X) \mathcal{N}(X|\mu,\sigma)\, dX = \frac{1}{N} \sum_{i=1}^N f(X_i) \mathcal{N}(X_i | \mu, \sigma),
\end{align}
where $X$ is drawn from a uniform distribution over all space, with 
\begin{align}
\int f(X) \mathcal{N}(X|\mu,\sigma)\, dX = \frac{1}{N} \sum_{i=1}^N f(X_i),
\end{align}
where $X$ drawn from $\mathcal{N}(X_i | \mu, \sigma)$.
